### 2014-02-26 thesis-proposal/att (Stheno's conflicted copy 2014-02-23).tex
More progress on Sec. 5

\section{Active Type Synthesis}\label{att}
General-purpose abstraction mechanisms, like those available in Wyvern and other languages today, are powerful, and should be used when possible. But relying on a small fixed collection of them may not always be satisfactory, for several reasons:
\begin{enumerate}
\item General-purpose abstraction mechanisms themselves are considerably varied. Tyconsider, for example, the many different ways languages expose product types: simple products, $n$-ary products, labeled products\footnote{We use the phrase \emph{labeled product} to avoid the connotations of related terms like \emph{record}, \emph{struct} and \emph{object}.}, structurally-typed labeled products, labeled products with mutable fields, labeled products with methods, labeled products with prototypic inheritance, labeled products with delegation, labeled products with class-based inheritance, and various combinations of these. New mechanisms and variants are constantly being developed. To allow these new general-purpose mechanisms to be evaluated in a controlled manner in practice, it must be possible to import them into existing codebases in a piecemeal fashion.
\item Specialized type systems that directly enforce stronger invariants than general-purpose abstraction mechanisms are capable of straightforwardly enforcing are often developed. In addition to the examples of type systems for regular expressions in Sec. \ref{regex} (\cite{a,b}), there are also specialized type systems for parallel programming \cite{a,b}, concurrent programming \cite{cml,a,b}, distributed programming \cite{tom7}, reactive programming \cite{reactiveml}, authenticated data structures \cite{popl13}, security policies \cite{walker00}, information flow \cite{smith2001}, web programming \cite{sandholm00}, aliasing \cite{naden12}, network protocols \cite{sekar99}, units of measure \cite{keneddy} and many others. These type systems are implemented by language-external means (e.g. by creating a new variant of ML, Java or C), leading to the problems of Sec. \ref{external-approaches}. 
\item Interoperability layers with external and legacy systems are another major area where support for directly implementing a new type system (that of the external language) and implementing its primitives in terms of a low-level interoperability API would be useful (e.g. \cite{CinJava}).
%\item General-purpose abstraction mechanisms are typically implemented uniformly. For example, inductive datatypes are typically implemented using tagged pointers to a payload. In some cases (e.g. natural numbers) it may be desirable to implement an abstraction in an alternative manner (e.g. as an integer) so that certain operations perform better. Parallel abstractions are of particular note, as they can benefit substantially from specialized implementation strategies on specialized hardware (e.g. a GPU).  Note that if control over internal implementation details were the only relevant use case, abstract datatypes might be appropriate.
\end{enumerate}

In this section, we will focus on language-integrated, type-oriented mechanisms for implementing extensions to the static and dynamic semantics of programming languages. We will leave the syntax fixed, inverting the setup in the previous section. We will begin by taking a  first-principles type-theoretic approach in Sec. \ref{foundations}, focusing on safety issues, and continue on in Sec. \ref{ace} by designing a full language called Ace. We will show how a range of statically-typed general-purpose abstraction mechanisms as well as more specialized abstractions can be expressed as safely-composable extensions within Ace. Notably, Ace is itself implemented as a library for Python. We discuss how to bootstrap a highly-expressive extensible static type system atop the quotation and object-oriented features available in Python, a uni-typed language.

\subsection{Foundations}\label{foundations}
Programming languages are typically organized around a monolithic collection of indexed type and operator constructors. Let us begin by considering the simply-typed lambda calculus (STLC). It provides a single type constructor, indexed by a pair of types, that we will call $\fvar{Arrow}$ for uniformity. It also provides two operator constructors: $\lambda$, indexed by a type, and $\texttt{ap}$, indexed trivially. We can specify the static semantics of the STLC as follows, using a uniform abstract syntax where type and operator indices are written in braces to emphasize this way of thinking about its structure:
\begin{figure}[h]
\small
\vspace{-8pt}
\begin{mathpar}
\inferrule[var]{ }{
	\jet{\eCtxX{\eCtx}{x}{\tau}}{x}{\tau}
}

\inferrule[arrow-I]{
	\jet{\eCtxX{\eCtx}{x}{\tau}}{e}{\tau'}	
}{
	\jetX{\lambda[\tau]({x}.e)}{\fvar{Arrow}[(\tau, \tau')]}
}

\inferrule[arrow-E]{
	\jetX{e_1}{\fvar{Arrow}[(\tau, \tau')]}\\
	\jetX{e_2}{\tau}
}{
	\jetX{\texttt{ap}[()](e_1; e_2)}{\tau'}
}
\end{mathpar}
\vspace{-8pt}
\caption{Static semantics of the simply typed lambda calculus.}
\end{figure}

Although a researcher may casually speak of ``conservatively extending the STLC`` with a primitive natural number type and its corresponding operators (as in G\"odel's T; Figure \ref{nat-statics}) or primitive product types and their  corresponding operators (Figure \ref{pair-statics}), it is clearly not possible to directly introduce these from within the STLC.
\begin{figure}[h]
\small
\vspace{-8pt}
\begin{mathpar}
\inferrule[nat-I1]{ }{
	\jetX{\texttt{z}[()]()}{\fvar{Nat}[()]}
}

\inferrule[nat-I2]{
	\jetX{e}{\fvar{Nat}[()]}
}{
	\jetX{\texttt{s}[()](e)}{\fvar{Nat}[()]}
}

\inferrule[nat-E]{
	\jetX{e_1}{\fvar{Nat}[()]}\\
	\jetX{e_2}{\tau}\\
	\jet{\eCtxX{\eCtxX{\eCtx}{x}{\fvar{Nat}[()]}}{y}{\tau}}{e_3}{\tau}
}{
	\jetX{\texttt{natrec}[()](e_1; e_2; x.y.e_3)}{\tau}
}
\end{mathpar}
\vspace{-8pt}
\caption{Static semantics of natural numbers.}\label{nat-statics}
\vspace{-15pt}
\end{figure}
\begin{figure}[h]
\small
\begin{mathpar}
\inferrule[Prod-I]{
	\jetX{e_1}{\tau_1}\\
	\jetX{e_2}{\tau_2}
}{
	\jetX{\texttt{pair}[()](e_1; e_2)}{\fvar{Prod}[(\tau_1, \tau_2)]}
}

\inferrule[Prod-E1]{
	\jetX{e}{\fvar{Prod}[(\tau_1, \tau_2)]}
}{
	\jetX{\texttt{prl}[()](e)}{\tau_1}
}

\inferrule[Prod-E2]{
	\jetX{e}{\fvar{Prod}[(\tau_1, \tau_2)]}
}{
	\jetX{\texttt{prr}[()](e)}{\tau_1}
}
\end{mathpar}
\vspace{-8pt}
\caption{Static semantics of products}\label{pair-statics}
\end{figure}

One recourse researchers have in such situations is to attempt to encode new constructs in terms of existing constructs. Collections of such encodings are sometimes called \emph{embedded domain-specific languages (EDSLs)} \cite{fowler2010domain}. Unfortunately, finding an encoding that fully captures both the static and dynamic semantics of a desirable construct can be difficult\footnote{For example, Church encodings in System F were among the more challenging topics for students in our undergraduate programming languages course, 15-312.} and is not always possible. Even when it possible, it may be impractical to work with and optimize. For our example of adding natural numbers or products to the STLC, a reasonable encoding is impossible. In closely-related languages, like the polymorphic lambda calculus (a.k.a. System F), such constructs are only weakly definable via Church encodings \cite{reynoldsIntroToPolymorphism} and far from practical. Reynolds, echoing Perlis   \cite{perlisPearl}, puts it succinctly \cite{reynoldsIntroToPolymorphism}:

\begin{quote}
To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program. It is clear that the language requires a novel programming style.
Moreover, it is likely that certain important functions cannot be expressed by their most
efficient algorithms.
\end{quote}

@@

An extensible programming language could address these problems by providing a language-integrated mechanism for introducing new type and operator constructors and implementing their associated static and dynamic semantics directly. 
%Library developers need only consider which abstractions are most appropriate for their domain, without also considering whether these constructs can be exposed using abstractions appropriate to the domains of client code. Clients can simply import any necessary constructs when using a library that relies on them, preserving safety and ease-of-use without the use of  wrappers and glue code. We show this competing approach in Figure \ref{approaches}(b).
%Researchers and domain experts thus gain the ability to distribute new ideas for evaluation to a broader development community without requiring the approval of maintainers of mainstream languages, large-scale porting of code or explicit interoperability layers. 
But, as mentioned in Section \ref{language-integrated-approaches}, some significant challenges must be addressed before such a mechanism can be relied upon. The desire for expressiveness must be balanced against  concerns about maintaining various safety properties in the presence of arbitrary combinations of user-defined  extensions to the language's core semantics. The mechanism must ensure that desirable \emph{metatheoretic properties} (e.g. type safety, decidability) of the language are maintained by extensions. Because multiple independently developed extensions might be used within one program, the mechanism must further guarantee their \emph{non-interference}. These are the issues we seek to address in this work.
%\subsection{Theory}\label{atlam}
%%\begin{figure}
%%\begin{mathpar}
%%\inferrule{a}{b}
%%\end{mathpar}
%%\end{figure}
%In this section, we will describe a minimal calculus that captures our language-internal extensibility mechanism,  called \emph{active typechecking and translation (AT\&T)}. AT\&T allows developers to declare new primitive type families, associate operators with them, implement their static semantics in a functional style, and realize their dynamic semantics by simultaneously implementing a translation into a typed internal language. Note that this latter mechanism is closely related to how Standard ML was (re-)specified\todo{cite/read a bit about this/ask Bob/flesh this out}, but that we are fundamentally interested in extending language \emph{implementations}, not their  declarative specifications; proving the adequacy of such implementations against mechanized specifications, or extracting them directly from such specifications, will be investigated in future work.
%
%The AT\&T mechanism utilizes type-level computation of higher kind and integrates typed compilation techniques into the language to allow us to give strong metatheoretic guarantees,  and uses a mechanism notionally related to abstract types (such as those found in the ML module system) to guarantee that extensions cannot interfere with one another,  while remaining straightforward and expressive. In this section, we will develop a core calculus, called \atlam, which uses a minimal, uniform grammar for primitive operators. Then in Section \ref{ace}, we will show how to realize this minimal mechanism within a widely-used language with a more expressive grammar.
%
%AT\&T is general with respect to many choices about the type-level language, the typed internal language and syntax. Choices along these dimensions can affect both expressiveness and ease-of-use. We will begin in Sec. 2 by introducing a minimal system called $@\lambda$ (the ``actively-typed lambda calculus'') that distills the essence of the mechanism in a simply-typed, simply-kinded setting. This will allow us to fully and concisely formalize the language and compiler and give several key safety theorems. We will then continue in Sec. 3 by discussing variants of this mechanism based on other basic paradigms, considering dependently-typed functional languages and object-oriented languages, discussing trade-offs between expressivity and safety when doing so. We have developed a simple prototype called Ace and have used it to develop a number of full-scale language extensions as libraries. We will briefly discuss this language and these extensions in Sec. 4.

@@

\subsubsection{From Extensible Compilers to Extensible Languages}\label{evolution}
The monolithic character of most programming languages is reflected in, and perhaps influenced by, the most common mechanisms used for implementing programming languages.
Let us consider an implementation of the STLC. 
A compiler written using a functional language will invariably represent the primitive type  and operator constructors using {closed} recursive sums. 
A simple implementation in Standard ML could be based around these datatypes, for example:
\begin{lstlisting}
  datatype Type = Arrow of Type * Type
  datatype Exp = Var of var | Lam of var * Type * Exp | Ap of Exp * Exp 
\end{lstlisting}

The compiler front-end, which consists of a typechecker and translator to a suitable intermediate language (for subsequent compilation by some suitable back-end), will proceed by exhaustive case analysis over the constructors of \lstinline{Exp}.

In a class-based object-oriented implementation of Godel's T, we might instead encode type and operator constructors as subclasses of abstract classes \lstinline{Type} and \lstinline{Exp}. Typechecking and translation, however, will proceed by the ubiquitous \emph{visitor pattern}, dispatching against a fixed collection of {known} subclasses of \lstinline{Exp}.

In either case, we encounter the same basic issue: there is no way to modularly add new primitive type and operator constructors to \li{Type} and \li{Exp}, respectively, and provide implementations of their associated typechecking and translation logic. Everything is defined in precisely one place.
%This issue is related to the widely-discussed \emph{expression problem} (in a restricted sense -- we do not consider adding new functions beyond typechecking and translation here, only adding logic to these) \cite{wadler-expression}.

A number of language mechanisms have been proposed that allow new cases to be added to datatypes and the functions that operate over them in a modular manner. 
In functional languages, we might use \emph{open datatypes} \cite{opendatatypes}\todo{cite}. For example, if we wish to extend our implementation of the STLC with product types and we have written our compiler in a language supporting open datatypes, it might be possible to add new constructors: 
\begin{lstlisting}
  newcon Prod of Type * Type extends Type
  newcon Pair of Exp * Exp extends Exp
  newcon PrL of Exp extends Exp
  newcon PrR of Exp extends Exp
\end{lstlisting}

The logic for functionality like typechecking and translation can then be implemented for only these new cases. For example, the \lstinline{typeof} function that synthesizes a type for an expression given a context could be extended to support the new case \li{PrL} like so:
\begin{lstlisting}
  typeof (ctx, PrL(e)) = case typeof (ctx, e) of 
      Prod(t1, _) => t1 
    | _ => raise TypeError("<appropriate error message>")
\end{lstlisting}

If we allowed users to define new modules containing definitions like these and link them into our compiler, we will have succeeded in creating an externally-extensible compiler, albeit one where safety and non-interference is not guaranteed (we will return to this point shortly). We have not, however, created an extensible programming language because other compilers for the same language will not necessarily support the same extensions. 
If our newly-introduced constructs are exposed at a library's  interface boundary, clients of that library who are using different compilers face the same problems with client compatibility that those using different languages face (as described in Sec. \ref{external-approaches}). That is, {extending a language by extending a single compiler for it is morally equivalent to creating a new language}.

We argue that a more appropriate and useful place for extensions like this is directly within libraries, alongside abstractions that can be implemented in terms of existing primitive abstractions. To enable this, the language must allow for the introduction of new type constructors, like \lstinline{Prod}, operator constructors, like \lstinline{Pair}, \lstinline{PrL} and \lstinline{PrR}, and their associated type synthesis and translation logic. Because this mechanism is integrated into the language specification, all compilers must support it.

The design described above suggests we may now need to add another layer to our language, above the type-level language (conventionally, $\tau$) and expression language (conventionally, $e$), where extensions are implemented. In fact, we will show that \textbf{a natural place for type system extensions is within the type-level language}. The intuition  is that extensions to a statically typed language's semantics will need to manipulate types as values at compile-time. Many languages already allow users to write type-level functions for various reasons, effectively supporting this notion of types as values at compile-time. The type-level language is often constrained by its own type system, where the types of type-level terms are called \emph{kinds} for clarity,  that prevents type-level functions from causing problems during compilation. The kind system is often more constraining than the type system is (e.g. it might permit only terminating functions, to ensure that typechecking is decidable). This is precisely the structure that a distinct extension layer would have, and so we will show that is quite natural to unify the two.

\subsubsection{@$\lambda$: Conservatively Extending a Type System From Within}
\begin{figure}[t]
\small
$$\begin{array}{rccl}	
\textbf{programs} & \rho & ::= & \pfam{\familyDf}{\progsort} \pipe \pdef{t}{\kappa}{\tau}{\progsort} \pipe e\\
\text{operators}		&	\theta	&	::= &	\tops{op}{\kappaidx}{i}{a}{\taut{def}} \pipe 
												\topp{\theta}{\theta}\\
\\
\textbf{external terms} 				&	e	&	::=	&	\evar{x} \pipe 
														\elam{\evar{x}}{\tau}{e} \pipe 
														\eop{Tycon}{op}{
															\tauidx
														}{
  												    		\splat{e}{1}{n}
														} \\
									& 		&		& 	\\

@@

>>>>

\section{Type-Oriented Mechanisms for Semantic Extensions}\label{att}

In this and the following two sections, we will focus on language-integrated, type-oriented mechanisms for implementing extensions to the static and dynamic semantics of programming languages. We will leave the syntax fixed, inverting the setup in the previous section. We do not believe  the two approaches to be incompatible with one another, but it is useful to consider the issues separately.

\subsection{Foundations}
Typed programming languages can be considered in terms of the collection of indexed type and operator constructors they provide. The simply-typed lambda calculus (STLC), for example, provides a single type constructor, \fvar{Arrow}, indexed by a pair of types, and two operator constructors: $\opvar{lam}$, indexed by a type, and $\opvar{ap}$, which can be thought of as being indexed trivially. It's static semantics can be written abstractly like this, to emphasize this uniform way of thinking about type structure:
\begin{figure}[h]
\small
\vspace{-8pt}
\begin{mathpar}
\inferrule[var]{ }{
	\jet{\eCtxX{\eCtx}{x}{\tau}}{x}{\tau}
}

\inferrule[arrow-I]{
	\jet{\eCtxX{\eCtx}{x}{\tau}}{e}{\tau'}	
}{
	\jetX{\opvar{lam}[\tau]({x}.e)}{\fvar{Arrow}[(\tau, \tau')]}
}

\inferrule[arrow-E]{
	\jetX{e_1}{\fvar{Arrow}[(\tau, \tau')]}\\
	\jetX{e_2}{\tau}
}{
	\jetX{\opvar{ap}[()](e_1; e_2)}{\tau'}
}
\end{mathpar}
\vspace{-8pt}
\caption{Static semantics of the simply typed lambda calculus.}
\end{figure}

We might extend this language with universally quantified and recursive types by adding the type constructors $\fvar{Forall}$ and $\fvar{Mu}$, which are indexed by types under a binder, and their associated operator constructors (see \cite{pfpl} for a review). It is tempting to stop here: the language now supports general recursion (i.e. it is  Turing complete) and the dynamic semantics of sums, products and inductive and co-inductive types can be defined by a method analagous to Church encodings in the untyped lambda calculus \cite{reynolds}\todo{what to cite for this?}. This is of clear theoretical and pedagogical interest, but this calculus is plainly impractical as a programming language. Reynolds, striking a cautionary tone reminiscent of Perlis in his description of the ``Turing tarpit'' \cite{Perl82a}, put it succinctly \cite{Reynolds94anintroduction}: 
\begin{quote}
To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program. It is clear that the language requires a novel programming style. Moreover, it is likely that certain important functions cannot be expressed by their most efficient algorithms.
\end{quote}

A simple calculus, equipped perhaps with some purely syntactic desugarings as given by a Church-style  weak encoding, can often suffice when only the dynamic semantics of abstractions are of importance. But continuing to build new type and operator constructors (collectively, \emph{constructs}) into a language can make statically reasoning about programs simpler and more precise, support a more natural programming style and endow the language with a more favorable cost semantics.

Consistent with this view, typed programming languages generally expose a richer set of constructs externally than their compilers work with internally.  
A functional language like ML, for example, might expose $n$-ary product types (tuples), labeled product types (records), recursive labeled sum types (datatypes) and forms of type abstraction (polymorphism and abstract  types). A compiler for the language might then, after typechecking a program according to the richer static semantics that these constructs define, translate it to an intermediate  language (IL) with only simple product and sum types, and perhaps some machine-oriented constructs like arrays, for further compilation to machine code. Correctly implementing the dynamic semantics of the language becomes the primary concern during this phase, so a simple IL that decreases the number of cases that must be considered when implementing and verifying the compiler is a wise choice.

It is again tempting to stop here. Indeed, languages like ML are often referred to as ``general-purpose languages'' and have been used successfully for a range of computing tasks. The constructs they provide certainly occupy a ``sweet spot'' in the design space, and a variety of higher-level abstractions can be strongly defined in a satisfying manner. But claims that they are suitable for \emph{all} purposes are not backed by any ``universality'' theorems stronger than those for the simple calculus described above. In fact, situations continue to arise in both research and practice where new constructs are clearly needed because a useful typed abstraction can only be weakly defined in terms of existing constructs, or the definition is impractically verbose or inefficient:

\begin{enumerate}
\item General-purpose abstractions continue to evolve and admit variants. There are many  variations on record types, for example: records with functional record update, prototypic delegation or fixed field ordering all have valid uses, and we will give all of these as examples. Similarly, sum types are awkward to encode and often unsafe to define using object-oriented constructs, and also admit variants (e.g. open sums, as we will discuss).

\item Perhaps more interestingly, specialized type systems that enforce stronger invariants than general-purpose constructs are capable of enforcing are often proposed by researchers. One need not take more than a cursory glance through the literature to discover specialized type systems for parallel programming \cite{a,b}, concurrent programming \cite{cml,a,b}, distributed programming \cite{tom7}, dataflow programming \cite{reactiveml}, authenticated data structures \cite{popl13}, information flow \cite{walker00, smith2001}, databases \cite{db}, aliased references \cite{naden12}, effects \cite{effects},  network protocols \cite{sekar99}, regular expressions \cite{regex}, units of measure \cite{keneddy} and many others. These examples are all implemented non-orthogonally (typically as extended versions of an existing language).
\item Interoperability layers that allow programmers to safely and naturally interact with libraries written in a different language require enforcing the type system of the foreign language within the calling language. Using  an interoperability layer that does not enforce these invariants in the calling language can lead to safety issues, even when both languages are separately known to be safe. Interoperability is a particularly important concern when transitioning to a new language, so this is likely a significant barrier to adoption of modern language constructs in practice.
\end{enumerate}
An extensible programming language could address these problems by providing a safe language-integrated mechanism for introducing new type and operator constructors and implementing their associated static and dynamic semantics directly, and we will explore this in the next two sections. Let us first take a detour into a typical compiler to better motivate the mechanisms that we will introduce into the language.
%Library developers need only consider which abstractions are most appropriate for their domain, without also considering whether these constructs can be exposed using abstractions appropriate to the domains of client code. Clients can simply import any necessary constructs when using a library that relies on them, preserving safety and ease-of-use without the use of  wrappers and glue code. We show this competing approach in Figure \ref{approaches}(b).
%Researchers and domain experts thus gain the ability to distribute new ideas for evaluation to a broader development community without requiring the approval of maintainers of mainstream languages, large-scale porting of code or explicit interoperability layers.

\subsection{From Extensible Compilers to Extensible Languages}\label{evolution}
The monolithic character of most programming languages is reflected in, and perhaps influenced by, the most common abstractions used for implementing programming languages.
Let us consider an implementation of the STLC. 
A compiler written using a functional language will invariably represent the primitive type  and operator constructors using {closed} recursive sums. 
For example, a simple implementation in Standard ML might be based around these datatypes (where variables, binders and operator application to arguments are handled separately, not shown)\footnote{This is essentially the approach that we took in the infrastructure for the coding assignments in 15-312.}:
\begin{lstlisting}
  datatype Ty = Arrow of Ty * Ty
  datatype Op = Lam of Ty | Ap 
\end{lstlisting}

%The compiler front-end, which consists of a typechecker and translator to a suitable intermediate language,    proceeds by exhaustive case analysis over the constructors of \verb|Op|.

In a class-based object-oriented implementation of the STLC, we might instead encode type and operator constructors as subclasses of abstract classes \verb|Ty| and \verb|Op|. If typechecking and translation proceed by the common \emph{visitor pattern}, dispatching against a fixed set of {known} subclasses of \verb|Op|, then we encounter a similar issue: everything is defined at once.

This issue was first discussed by Reynolds \cite{Reynolds75} and was dubbed the \emph{expression problem} by Wadler \cite{wadler1998expression}. A number of mechanisms have since been proposed that allow new cases to be added to data types and the functions that operate over them in a modular manner. 
In functional languages, we might turn to \emph{open datatypes and open functions} \cite{conf/ppdp/LohH06, Millstein:2002:MTH:581478.581489}. For example, we might add a new type constructor, \verb|Tuple|, indexed by a list of types, and two new operator constructors: an introductory form, \verb|NewTuple|, indexed trivially, and an elimination form, \verb|Prj|, indexed by a natural number, as follows:
\begin{lstlisting}
  newcon Tuple of Ty list extends Ty
  newcon NewTuple extends Op
  newcon Prj of nat extends Op
\end{lstlisting}

The logic for functionality like typechecking and translation, if written using open functions, can then be implemented for only these new cases. For example, the \verb|optype| function that synthesizes a type for an operator invocation given a context and a list of arguments might be extended to support the new constructor \verb|Prj| like so:
\begin{lstlisting}
  optype (ctx, Prj(n), [e]) = case exptype (ctx, e) of 
      Tuple(ts) => (case List.nth n ts of 
          Some t => t
        | _ => raise IndexError("<projection index out of bounds>"))
    | _ => raise TypeError("<tuple expected>")
  optype (ctx, Prj(n), _) = raise ArityError("<expected 1 argument>")
\end{lstlisting}

In a class-based object-oriented language like Java, we might dispense with the visitor pattern and instead invert control to methods of \verb|Op|.

\begin{lstlisting}[language=Java]
class Prj extends Op {
  Prj(Nat n) { this.n = n; }
  Nat n;
  Ty optype(Context ctx, List<Exp> args) {
    if (args.length != 1) throw new ArityError("<expected 1 argument>")
    Ty t = args[0].exptype(ctx);
    if (t instanceof Tuple) {
      Tuple t = (Tuple) t;
      try {
        return t.types[this.n]
      } catch OutOfBoundsException(e) {
        throw new IndexError("<projection index out of bounds>");
      }
    } else throw new TypeError("<tuple expected>");
  }
}
\end{lstlisting}

If we allowed users to load definitions like these into our compiler, we will have succeeded in creating an  extensible compiler, albeit one where orthogonality is not guaranteed (we will return to this point). We have not, however, created an extensible programming language because other compilers for the same language will not necessarily support the same extensions. These mechanisms are not \emph{language-integrated}, so if our newly-introduced constructs are exposed at a library's  interface boundary, clients of that library who are using different compilers face the same problems with client compatibility that those using different languages face (as described in Sec. \ref{external-approaches}). %That is, {extending a language by extending a single compiler for it is morally equivalent to creating a new language}. 
Nevertheless, these two strategies give us the conceptual seeds for the approaches that we will take in Secs. \ref{atlam} and \ref{ace}, respectively. 
In both cases, we introduce into the language a mechanism similar to, but more constrained and specialized  than, the two above.  Being part of the language, all compilers must then support all extensions, avoiding the client compatibility issue.\footnote{For  languages where a canonical implementation {is}, unfortunately, the specification (e.g. Haskell and Scala), the mechanisms we introduce can be left in the compiler if the build system loads and activates all needed extensions when linking two libraries. The advances related to safety, orthogonality and conservativity apply also to this design, but we do not consider it further in this thesis.}

% We will show how, by integrating typed compilation techniques and a form of data hiding with similarities to abstract types into the semantics of operators, we can achieve type safety and conservativity of the language as a whole, despite this level of extensibility. Decidability is achieved by ensuring that the type-level language is total. Type equality is made decidable in a simple manner: by preventing type-level functions from appearing in type indices.

%The design described above suggests we may now need to add another layer to our language, above the type-level language (conventionally, $\tau$) and expression language (conventionally, $e$), where extensions are implemented. In fact, we will show that \textbf{a natural place for type system extensions is within the type-level language}. The intuition  is that extensions to a statically typed language's semantics will need to manipulate types as values at compile-time. Many languages already allow users to write type-level functions for various reasons, effectively supporting this notion of types as values at compile-time. The type-level language is often constrained by its own type system, where the types of type-level terms are called \emph{kinds} for clarity,  that prevents type-level functions from causing problems during compilation. The kind system is often more constraining than the type system is (e.g. it might permit only terminating functions, to ensure that typechecking is decidable). This is precisely the structure that a distinct extension layer would have, and so we will show that is quite natural to unify the two.

We begin in Sec. \ref{atlam} by developing a core calculus called @$\lambda$ and focusing on theoretical issues, including type safety, decidability and {conservativity} (that extensions cannot weaken one another, in any combination). We will also sketch a flexible type-directed desugaring mechanism that targets this core calculus and give examples of simple but interesting extensions that are only weakly definable using the general-purpose constructs in functional languages. We then continue on in Sec. \ref{ace} by designing a full programming language called Ace. We will show richer examples of statically-typed general-purpose abstractions, as well as more specialized abstractions, that can be expressed as extensions within Ace. Ace  uses Python as its type-level language and supports the definition of user-defined internal languages, rather than picking a fixed target as in @$\lambda$.  Using a type-level language that provides few strictly enforced safety guarantees makes giving strong theoretical results about Ace more difficult, but we include checks similar to those in @$\lambda$ that require intention to subvert. The claimed contribution is thus fundamentally that Ace shows how to {bootstrap} a highly-expressive extensible static type system atop facilities, like quotations and objects, that are already available in widely-adopted languages. We argue that this is a useful bridge between existing monolithic languages and extensible languages. We conjecture that it would be possible to bootstrap a statically-typed subset of Python that, if used ubiquitously during compilation, would be sufficient to provide stronger theoretical guarantees as well.

%But, as mentioned in Section \ref{language-integrated-approaches}, some significant challenges must be addressed before such a mechanism can be relied upon. The desire for expressiveness must be balanced against  concerns about maintaining various safety properties in the presence of arbitrary combinations of user-defined  extensions to the language's core semantics. The mechanism must ensure that desirable \emph{metatheoretic properties} (e.g. type safety, decidability) of the language are maintained by extensions. Because multiple independently developed extensions might be used within one program, the mechanism must further guarantee that extensions are \emph{conservative}. These are the issues we seek to address in the next two sections.

\section{@$\lambda$}\label{atlam}
%Programming languages are typically organized around a monolithic collection of indexed type and operator constructors. Let us begin by considering the simply-typed lambda calculus (STLC). It provides a single type constructor, indexed by a pair of types, that we will call $\fvar{Arrow}$ for uniformity. It also provides two operator constructors: $\lambda$, indexed by a type, and $\texttt{ap}$, indexed trivially. We can specify the static semantics of the STLC as follows, using a uniform abstract syntax where type and operator indices are written in braces to emphasize this way of thinking about its structure:
%
%Although a researcher may casually speak of ``conservatively extending the STLC`` with a primitive natural number type and its corresponding operators (as in G\"odel's T; Figure \ref{nat-statics}) or primitive product types and their  corresponding operators (Figure \ref{pair-statics}), it is clearly not possible to directly introduce these from within the STLC.
%\begin{figure}[h]
%\small
%\vspace{-8pt}
%\begin{mathpar}
%\inferrule[nat-I1]{ }{
%	\jetX{\texttt{z}[()]()}{\fvar{Nat}[()]}
%}
%
%\inferrule[nat-I2]{
%	\jetX{e}{\fvar{Nat}[()]}
%}{
%	\jetX{\texttt{s}[()](e)}{\fvar{Nat}[()]}
%}
%
%\inferrule[nat-E]{
%	\jetX{e_1}{\fvar{Nat}[()]}\\
%	\jetX{e_2}{\tau}\\
%	\jet{\eCtxX{\eCtxX{\eCtx}{x}{\fvar{Nat}[()]}}{y}{\tau}}{e_3}{\tau}
%}{
%	\jetX{\texttt{natrec}[()](e_1; e_2; x.y.e_3)}{\tau}
%}
%\end{mathpar}
%\vspace{-8pt}
%\caption{Static semantics of natural numbers.}\label{nat-statics}
%\vspace{-15pt}
%\end{figure}
%\begin{figure}[h]
%\small
%\begin{mathpar}
%\inferrule[Prod-I]{
%	\jetX{e_1}{\tau_1}\\
%	\jetX{e_2}{\tau_2}
%}{
%	\jetX{\texttt{pair}[()](e_1; e_2)}{\fvar{Prod}[(\tau_1, \tau_2)]}
%}
%
%\inferrule[Prod-E1]{
%	\jetX{e}{\fvar{Prod}[(\tau_1, \tau_2)]}
%}{
%	\jetX{\texttt{prl}[()](e)}{\tau_1}
%}
%
%\inferrule[Prod-E2]{
%	\jetX{e}{\fvar{Prod}[(\tau_1, \tau_2)]}
%}{
%	\jetX{\texttt{prr}[()](e)}{\tau_1}
%}
%\end{mathpar}
%\vspace{-8pt}
%\caption{Static semantics of products}\label{pair-statics}
%\end{figure}
%
%One recourse researchers have in such situations is to attempt to encode new constructs in terms of existing constructs. Collections of such encodings are sometimes called \emph{embedded domain-specific languages (EDSLs)} \cite{fowler2010domain}. Unfortunately, finding an encoding that fully captures both the static and dynamic semantics of a desirable construct can be difficult\footnote{For example, Church encodings in System F were among the more challenging topics for students in our undergraduate programming languages course, 15-312.} and is not always possible. Even when it possible, it may be impractical to work with and optimize. For our example of adding natural numbers or products to the STLC, an encoding is impossible. In closely-related languages, like the polymorphic lambda calculus (a.k.a. System F), such constructs are only weakly definable via Church encodings \cite{reynoldsIntroToPolymorphism} and far from practical. Reynolds, echoing Perlis   \cite{perlisPearl}, puts it succinctly \cite{reynoldsIntroToPolymorphism}:
%
%\begin{quote}
%To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program. It is clear that the language requires a novel programming style.
%Moreover, it is likely that certain important functions cannot be expressed by their most
%efficient algorithms.
%\end{quote}
%

@@

%\subsection{Theory}\label{atlam}
%%\begin{figure}
%%\begin{mathpar}
%%\inferrule{a}{b}
%%\end{mathpar}
%%\end{figure}
%In this section, we will describe a minimal calculus that captures our language-internal extensibility mechanism,  called \emph{active typechecking and translation (AT\&T)}. AT\&T allows developers to declare new primitive type families, associate operators with them, implement their static semantics in a functional style, and realize their dynamic semantics by simultaneously implementing a translation into a typed internal language. Note that this latter mechanism is closely related to how Standard ML was (re-)specified\todo{cite/read a bit about this/ask Bob/flesh this out}, but that we are fundamentally interested in extending language \emph{implementations}, not their  declarative specifications; proving the adequacy of such implementations against mechanized specifications, or extracting them directly from such specifications, will be investigated in future work.
%
%The AT\&T mechanism utilizes type-level computation of higher kind and integrates typed compilation techniques into the language to allow us to give strong metatheoretic guarantees,  and uses a mechanism notionally related to abstract types (such as those found in the ML module system) to guarantee that extensions cannot interfere with one another,  while remaining straightforward and expressive. In this section, we will develop a core calculus, called \atlam, which uses a minimal, uniform grammar for primitive operators. Then in Section \ref{ace}, we will show how to realize this minimal mechanism within a widely-used language with a more expressive grammar.
%
%AT\&T is general with respect to many choices about the type-level language, the typed internal language and syntax. Choices along these dimensions can affect both expressiveness and ease-of-use. We will begin in Sec. 2 by introducing a minimal system called $@\lambda$ (the ``actively-typed lambda calculus'') that distills the essence of the mechanism in a simply-typed, simply-kinded setting. This will allow us to fully and concisely formalize the language and compiler and give several key safety theorems. We will then continue in Sec. 3 by discussing variants of this mechanism based on other basic paradigms, considering dependently-typed functional languages and object-oriented languages, discussing trade-offs between expressivity and safety when doing so. We have developed a simple prototype called Ace and have used it to develop a number of full-scale language extensions as libraries. We will briefly discuss this language and these extensions in Sec. 4.

@@

%\subsubsection{@$\lambda$: Conservatively Extending a Type System From Within}
\begin{figure}[t]
\small
$$\begin{array}{rccl}	
\textbf{programs} & \rho & ::= & \pfam{\familyDf}{\progsort} \pipe \pdef{t}{\kappa}{\tau}{\progsort} \pipe e\\
		&	\theta	&	::= &	\tops{op}{\kappaidx}{i}{a}{\taut{def}} \pipe 
												\topp{\theta}{\theta}\\
\\
\textbf{external terms} 				&	e	&	::=	&	\evar{x} \pipe 
														\elam{\evar{x}}{\tau}{e} \pipe 
														\eop{Tycon}{op}{
															\tauidx
														}{
  												    		\splat{e}{1}{n}
														} \\
									& 		&		& 	\\

@@

--
